Deciphering enhancer rules underlying early human neuronal and neural crest development.

### 2025-03-25 manuscript.tex
Update references.

To investigate the enhancer code underlying dorso-ventral patterning of neuronal progenitors further, we calculated
contribution scores on the top 1,000 regions of each organoid and embryo dorso-ventral topic using respectively $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$. We then used TF-MoDISco\cite{shrikumar2020tfmodisco} to find reoccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities (SUPPLEMENTAL FIG). Next, we calculated the average contribution for all identified instances per cluster and for each topic resulting in a dorso-ventral enhancer code table (\textbf{Fig. 3e}). Both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns
underlying the different dorso-ventral states and the importance of the instances of these patterns is highly correlated across both
models (CORR COEF; \textbf{Fig. 3e}). For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD
and RFX instances is important (\textbf{Fig. 3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA2}, \textit{TEAD1} and \textit{RFX4} in both the organoid and embryo cells (\textbf{Fig. 3g}). In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-2} expression (\textbf{(Fig. 3e- g)}); PAX instances, correlating with \textit{PAX6} expression (\textbf{Fig. 3e-g}) and ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3e-g}). For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), SOX instances are important which correlates with \textit{SOX2} expression. Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3e}). Finally, we calculated the Jaccard index of genomic regions based on the presence of an instance of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig. 3h}). FOXA2, TEAD1 and RFX4 instances often co-occur in the same genomic region while instances of other patterns seem to be present in separate regions (\textbf{Fig. 3h}).\par

It has been reported that FOXA2 is a pioneer factor. Indeed, when centering on FOXA2 instances in both the organoid and
embryo data we observe a strong footprint when looking at Tn5 cut-sites (\textbf{Fig. 3i} and SUPPLEMENTAL FIG). We then
asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this
TF and regions, bound by FOXA2, that are specific to the floorplate. For this purpose, we downloaded FOXA2 ChIP-seq
peaks in HepG2 and A549 and intersected these with floorplate peaks containing FOXA2 instances. Those floorplate peaks
that overlapped with both ChIP-seq datasets we classified as regions that are generally bound by FOXA2 and those that
did not we classified as floorplate specific. Next, we quantified both the number of FOXA2 instances per region and the
affinity of each individual instance (making use of the FOXA2 motif score calculated using
FIMO\cite{grant2011fimo,schreiber2025tangermeme}) hypothesizing that the overall affinity of regions that are generally
bound by FOXA2 is higher than floorplate specific regions either because these general regions have more instances or
the affinity of individual instances is higher. Indeed, FOXA2 target regions that are specific to the floorplate mostly
have a single FOXA2 instance while FOXA2 target regions that are general tend to have more instances (\textbf{Fig. 3j}).
Furthermore, in case there is but a single FOXA2 instance in a general FOXA2 target the affinity of this instance tends
to be higher (\textbf{Fig. 3k}). Generating a \textit{de novo} motif from general and specific FOXA2 target regions that
have a single instance, we observe that FOXA2 instances in regions that are general tend to be of the form
\textit{TGTTT\underline{AC}} while those instances in specific regions tend to be of the form \textit{TGTTT} (\textbf{Fig. 3l}).\par

@@

\subsection*{
  Figure 3. Deciphering the enhancer code underlying dorso-ventral patterning of neuronal progenitors.
}
\textbf{a}, UMAP of organoid (top) and embryo (bottom) neuronal progenitors colored by cell-topic probabilities in RGB color scales of dorso-ventral topics. \textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-ventral topics ($n_{topic 33}$ = X, $n_{topic 38}$ = X, $n_{topic 36}$ = X, $n_{topic 54}$ = , and $n_{topic 48}$ = for organoid cells; and $n_{topic 34}$ = X, $n_{topic 38}$ = X, $n_{topic 79}$ = X, $n_{topic 88}$ = , and $n_{topic 58}$ = for embryo cells). \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr:...-...; hg38. SFPE2 and SFPE1 is highlighted. \textbf{d}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). \textbf{e}, Code table showing average contribution score of pattern instances for each dorso-ventral topic. Contribution scores for $DeepNeuralTube_{organoid}$ are not outlined. Contribution scores for $DeepNeuralTube_{embryo}$ are outlined and negated. \textbf{f}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics. \textbf{g}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics. \textbf{h}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. \textbf{i}, Scaled average and normalized number of cut sites (black, top) and contribution score (red, top) across genomic regions with a FOXA2 instance and heatmap showing normalized number of cut sites (bottom) for individual regions with a FOXA2 instance. Signal is centered on the FOXA2 instance start position. \textbf{j}, Bar chart showing the fraction of genomic regions that has a given number of FOXA2 instances for FOXA2 regions that overlap with HepG2 or A549 FOXA2 ChIp-seq signal (general; bottom) and those regions that don't overlap (specific; top). \textbf{k}, Boxplot (ADD NUMEBRS) showing $-log_10(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grant2011fimo,schreiber2025tangermeme} for general regions (bottom) and specific regions (top) stratified by the number of FOXA2 instance per region. \textbf{l}, \textit{de novo} motif for regions with a single FOXA2 instance that are specific (left, top) or general (right, top) and heatmap color coded by nucleotide identity for individual regions for both specific (left, bottom) and general (right, bottom) regions that have a single FOXA2 instance.

@@

Focusing on the two most anterior domains, two similar homeodomain TF binding motifs seem to underlie the differential
accessibility of those domains. Both have the core homeobox motif: \textit{TAAT} but one is flanked by the nucleotides \textit{TAG} while the other is flanked by the nucleotides \textit{CCCT}. We were surprised that this slight difference in the two motifs would be enough for encoding two separate AP domains. To test this hypothesis, we first did a substitution experiment where we replaced all instances of \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} in regions of the most anterior domain and vice versa (\textbf{Fig. 4f}). Indeed, $DeepNeuralTube_{embryo}$ predicts an anterior-to-posterior shift of accessibility of those regions after the \textit{CTAAT\underline{TAG}} to \textit{CTAAT\underline{CCCT}} substitution and vice versa a posterior-to-anterior shift of accessibility after the \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} to  \textit{CTAAT\underline{TAG}} substitution (\textbf{Fig. 4f}). Next, we designed enhancers from scratch by embedding either the  \textit{CTAAT\underline{TAG}} or \textit{CTAAT\underline{CCCT}} motif\cite{taskiran2024synthetic, kempynck2025crested} and using respectively Topic 61 and Topic 31 as target class (\textbf{Fig. 4g}). [SWITCH PANELS]. Embedding of two instances of either motif was sufficient to reach predicted chromatin accessibility levels of equivalent ranges to the predicted chromatin accessibility levels of genomic regions (\textbf{Fig. 4g}). Next, to eliminate potential model bias we repeated the same experiment but optimized for the opposite class. This is, we optimized for Topic 31 by embedding the \textit{CTAAT\underline{TAG}} motif optimized for Topic 61 by embedding the \textit{CTAAT\underline{CCCT}} motif (\textbf{Fig. 4h}). In this case, the synthetic design failed to converge for the target class but instead regions with high prediction score for the AP domain that corresponds to the motif used for embedding were generated. This is embedding of the \textit{CTAAT\underline{TAG}} motif still generated regions with high prediction score for Topic 61 and embedding of the \textit{CTAAT\underline{CCCT}} motif generated regions with high prediction score for Topic 31 (\textbf{Fig. 4h}).\par

@@

Organoid Topic 62 and embryo Topic 103 represent the pre-migratory neural crest state and instances of TEAD, GRHL, AP2 and ZEB patterns are important for this state (\textbf{Fig. 5b-c}) the importance of TEAD instances correlates with \textit{TEAD4} expression, GRHL with both \textit{GRHL1} and \textit{GRHL2} and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig. 5d}). Note that ZEB instances have a positive contribution to the pre-migratory while it has a negative contribution to all other neural crest states (\textbf{Fig. 5b-c}) and moreover all other states of both the neuronal progenitors (\textbf{Fig. 3e}) and neurons (\textbf{Fig. 6b}). The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig. 5d}). This is in line with the previously reported role of ZEB2 as a chromatin closing repressor \cite{taskiran2024synthetic}.\par

The migratory neural crest cell state is represented by two organoid topics (Topic 65 and 59) and a single embryo topic (Topic 94). Nuclear receptor, AP2, ZIC (only in organoid topics) and SOX-dimer (organoid Topic 59 but not 65) are important for this state (\textbf{Fig. 5b-c}). We could not link the nuclear receptor instances to a specific TF based on the correlation of expression and the importance score of these instances. The importance of AP2 instances correlates with the expression of \textit{TFAP2A}, \textit{TFAP2B} and \textit{TFAP2C} (\textbf{Fig. 5d}), ZIC with \textit{ZIC1} to \textit{ZIC5} and the SOX-dimer with expression of \textit{SOX10}. The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothstein2020heterodmerization}. After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) for which instances of RFX are important (\textbf{Fig. 5b-c}) that correlate with the expression of \textit{RFX4} (\textbf{Fig. 5d}).\par

Finally, the facial mesenchyme cell state is represented by organoid Topic 58 and embryo Topic 91. Instances of a FOX
pattern and a dimer consisting of a homeobox and ebox motif are important for this state (\textbf{Fig. 5b}). The
importance of the FOX instances correlates with the expression of \textit{FOXC2} and that of the dimer motif with the
expression of both \textit{TWIST1} and \textit{ALX4}. The latter is consistent with a recent study where it has been
shown that TWIST1 and ALX4 cooperatively bind this specific dimer motif regulating the expression of genes involved in
facial mesenchyme\cite{kim2024dnaguided}.\par

@@

Recently, a genome wide association study (GWAS) was published identifying 203 single-nucleotide polymorphisms (SNPs)
that are associated with facial variation\cite{white2021insights}. Given that regions surrounding these SNPs were
enriched for enhancer activity in cranial neural crest cells\cite{white2021insights} we asked if, using
$DeepNeuralTube$, we could explain the effect of some of these variants on enhancer activity. Therefore for all
suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ and calculated the difference in prediction score across both alleles (delta prediction score). At an absolute delta prediction score threshold of 0.5 we identified X SNPs that we deem explainable by either of the two models. Note that only very few (X out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig. 5f}). Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (SUPPLEMENTAL FIG). As an example we show the explanation of rs1555067 which is located upstream of \textit{PRRX1}. The G>A substitution of this SNP causes a delta prediction score of 0.X for organoid Topic 58 (\textbf{Fig. 5g}), increasing the predicted chromatin accessibility of this genomic region in the facial mesenchyme cell state. This predicted increase in chromatin accessibility is caused by generating an EBOX site 6bp upsteam of a partial homeobox site potentially creating a new binding site for TWIST1-ALX4 dimers (\textbf{Fig. 5h}).\par

@@

\subsection*{
  Figure 5. Deciphering the enhancer code underlying pre-migratory and migratory neural crest and facial mesenchyme.
}
\textbf{a}, UMAP of organoid (top two) and embryo (bottom two) neural crest cells colored by cell-topic probabilities in RGB color scales for neural crest related topics. \textbf{b}, Code table showing average contribution score of pattern instances for each neural crest related topic. Contribution scores for $DeepNeuralTube_{organoid}$ are not outlined. Contribution scores for $DeepNeuralTube_{embryo}$ are outlined and negated. \textbf{c}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) of neural crest related topics. \textbf{d}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) for neural crest related topics. \textbf{e}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. \textbf{f}, Manhattan plot showing -log p value of the SNP being associated with variations in facial shape\cite{white2021insights} (positive y-axis) and the maximum (across classes of the model) of the absolute value of the difference in prediction score between the reference and alternative allele (negative y-axis; circles for $DeepNeuralTube_{organoid}$ and crosses for $DeepNeuralTube_{embryo}$) across the human genome. Lead SNPs are shown with a bigger dot size. \textbf{g}, Zoom in of chromosome one with the locus of SNP rs1555067 shown as an inset, the SNP itself is shown in red. \textbf{h}, Nucleotide contribution scores for Topic 58 of $DeepNeuralTube_{organoid}$ of the references (top) and alternative (bottom) allele of SNP rs1555067. The location of the SNP is indicated with a dotted line.

@@

Summarizing our results, we identified a total of sixteen and fifteen distinct cell states in respectively the neural
tube organoids and four weeks post conception embryo head with almost one-to-one correspondence based on TFBS importances and TF expression. Additionally, we identified 4 distinct cell states corresponding to AP patterning of the neuronal progenitors in the embryo. To get insight into the minimal set of TFBS that are needed to encode these cell states we trained logistic regression models to classify genomic regions as to belonging to the topic representing each state or not, making use of the motif scores of the identified patterns from each code table (\textbf{Fig. 3e}, \textbf{Fig. 4e}, \textbf{Fig. 5b} and \textbf{Fig. 6b}). The overall performance of these models was acceptably high (\textbf{Fig. 7a}) with an average area under the ROC of X and area under the precision-recall curve of X. Notably, all models for neural crest related states performed best (\textbf{Fig. 7a}). Next, we selected patterns that had a positive coefficient for at least one of the logistic regression models resulting in a total of twenty one patterns for fifteen to sixteen states plus the additional four AP states (\textbf{Fig. 7a}). Finally, we used the identified patterns to design synthetic enhancers for each state using motif embedding\cite{taskiran2024synthetic, kempynck2025crested}. We used at least three TFBS in the design process for each state. In case less than three distinct patterns had a positive coefficient for a state, we implanted multiple copies of each pattern to come to a total of at least four TFBS. The resulting synthetic DNA sequences had an overall high prediction score (X.X on average) (\textbf{Fig. 7a}) - and were cell type specific (\textbf{Fig. 7b}) - with the prediction score for neural crest related states being again the highest. Note that some TFBS like FOX and SOX are used by many cell states (\textbf{Fig. 7a}) while other are more specific to certain states. \par

@@

\subsection*{
  Figure 7. Neural tube cell states are encoded using four to five distinct transcription factor binding
  sites
}.
\textbf{a}, Area under the receiver-operator curve (outlined dots) and precision-recall curve of logistic regression
models classifying genomic regions to each topic using motif scores as features (top) and patterns with positive
coefficient for each logistic regression model along with the number of each patterns used for enhancer design
represented by the size of each dot (middle) and boxplots (n = 200) showing prediction score of designed enhancers for
each topic (bottom). \textbf{b}, heatmap showing prediction score for each topic of enhancers designed for all topics
for using $DeepNeuralTube_{organoid}$ (left) and $DeepNeuralTube_{embryo}$ (right). \textbf{c}, scatter plot showing
location of all pattern instances and contribution score at those locations, color is -log p value of the motif score
for each pattern calculated using FIMO\cite{grant2011fimo, schreiber2025tangermeme} for organoid (top left) and embryo
(top right) and histograms showing the number of motif instances per 25 bp bins with confidence intervals indicated in
gray scales (black is +/- 1 standard deviation, gray is +/- 2 standard deviation and white is more than 3 standard
deviations) for organoid (bottom left) and embryo (bottom right).

@@

\noindent{The following Cell Press journals require most research articles to include their methods in a \textbf{METHODS} section, which should appear after the discussion: \textit{Cell Biomaterials}, \textit{Chem}, \textit{Chem Catalysis}, \textit{Cell Reports Physical Science}, \textit{Cell Reports Sustainability}, \textit{Device}, \textit{Joule}, \textit{Matter}, \textit{Newton}, \textit{One Earth}, and \textit{Patterns}. For all other Cell Press journals, please \textbf{delete} the contents of this page from the template and refer instead to \textbf{STAR Methods} on page 9.}

\section*{METHODS}

>>>>

To investigate the enhancer code underlying dorso-ventral patterning of neuronal progenitors further, we calculated
contribution scores on the top 1,000 regions of each organoid and embryo dorso-ventral topic using respectively $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$. We then used TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018} to find reoccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities (SUPPLEMENTAL FIG). Next, we calculated the average contribution for all identified instances per cluster and for each topic resulting in a dorso-ventral enhancer code table (\textbf{Fig. 3e}). Both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns
underlying the different dorso-ventral states and the importance of the instances of these patterns is highly correlated across both
models (CORR COEF; \textbf{Fig. 3e}). For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD
and RFX instances is important (\textbf{Fig. 3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA2}, \textit{TEAD1} and \textit{RFX4} in both the organoid and embryo cells (\textbf{Fig. 3g}). In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-2} expression (\textbf{(Fig. 3e- g)}); PAX instances, correlating with \textit{PAX6} expression (\textbf{Fig. 3e-g}) and ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3e-g}). For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), SOX instances are important which correlates with \textit{SOX2} expression. Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3e}). Finally, we calculated the Jaccard index of genomic regions based on the presence of an instance of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig. 3h}). FOXA2, TEAD1 and RFX4 instances often co-occur in the same genomic region while instances of other patterns seem to be present in separate regions (\textbf{Fig. 3h}).\par

It has been reported that FOXA2 is a pioneer factor. Indeed, when centering on FOXA2 instances in both the organoid and
embryo data we observe a strong footprint when looking at Tn5 cut-sites (\textbf{Fig. 3i} and SUPPLEMENTAL FIG). We then
asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this
TF and regions, bound by FOXA2, that are specific to the floorplate. For this purpose, we downloaded FOXA2 ChIP-seq
peaks in HepG2 and A549 and intersected these with floorplate peaks containing FOXA2 instances. Those floorplate peaks
that overlapped with both ChIP-seq datasets we classified as regions that are generally bound by FOXA2 and those that
did not we classified as floorplate specific. Next, we quantified both the number of FOXA2 instances per region and the
affinity of each individual instance (making use of the FOXA2 motif score calculated using
FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) hypothesizing that the overall affinity of regions that are generally
bound by FOXA2 is higher than floorplate specific regions either because these general regions have more instances or
the affinity of individual instances is higher. Indeed, FOXA2 target regions that are specific to the floorplate mostly
have a single FOXA2 instance while FOXA2 target regions that are general tend to have more instances (\textbf{Fig. 3j}).
Furthermore, in case there is but a single FOXA2 instance in a general FOXA2 target the affinity of this instance tends
to be higher (\textbf{Fig. 3k}). Generating a \textit{de novo} motif from general and specific FOXA2 target regions that
have a single instance, we observe that FOXA2 instances in regions that are general tend to be of the form
\textit{TGTTT\underline{AC}} while those instances in specific regions tend to be of the form \textit{TGTTT} (\textbf{Fig. 3l}).\par

@@

\subsection*{
  Figure 3. Deciphering the enhancer code underlying dorso-ventral patterning of neuronal progenitors.
}
\textbf{a}, UMAP of organoid (top) and embryo (bottom) neuronal progenitors colored by cell-topic probabilities in RGB color scales of dorso-ventral topics. \textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-ventral topics ($n_{topic 33}$ = X, $n_{topic 38}$ = X, $n_{topic 36}$ = X, $n_{topic 54}$ = , and $n_{topic 48}$ = for organoid cells; and $n_{topic 34}$ = X, $n_{topic 38}$ = X, $n_{topic 79}$ = X, $n_{topic 88}$ = , and $n_{topic 58}$ = for embryo cells). \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr:...-...; hg38. SFPE2 and SFPE1 is highlighted. \textbf{d}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). \textbf{e}, Code table showing average contribution score of pattern instances for each dorso-ventral topic. Contribution scores for $DeepNeuralTube_{organoid}$ are not outlined. Contribution scores for $DeepNeuralTube_{embryo}$ are outlined and negated. \textbf{f}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics. \textbf{g}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics. \textbf{h}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. \textbf{i}, Scaled average and normalized number of cut sites (black, top) and contribution score (red, top) across genomic regions with a FOXA2 instance and heatmap showing normalized number of cut sites (bottom) for individual regions with a FOXA2 instance. Signal is centered on the FOXA2 instance start position. \textbf{j}, Bar chart showing the fraction of genomic regions that has a given number of FOXA2 instances for FOXA2 regions that overlap with HepG2 or A549 FOXA2 ChIp-seq signal (general; bottom) and those regions that don't overlap (specific; top). \textbf{k}, Boxplot (ADD NUMEBRS) showing $-log_10(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grantFIMOScanningOccurrences2011,schreiberTangermemeBiologicalSequence2025} for general regions (bottom) and specific regions (top) stratified by the number of FOXA2 instance per region. \textbf{l}, \textit{de novo} motif for regions with a single FOXA2 instance that are specific (left, top) or general (right, top) and heatmap color coded by nucleotide identity for individual regions for both specific (left, bottom) and general (right, bottom) regions that have a single FOXA2 instance.

@@

Focusing on the two most anterior domains, two similar homeodomain TF binding motifs seem to underlie the differential
accessibility of those domains. Both have the core homeobox motif: \textit{TAAT} but one is flanked by the nucleotides \textit{TAG} while the other is flanked by the nucleotides \textit{CCCT}. We were surprised that this slight difference in the two motifs would be enough for encoding two separate AP domains. To test this hypothesis, we first did a substitution experiment where we replaced all instances of \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} in regions of the most anterior domain and vice versa (\textbf{Fig. 4f}). Indeed, $DeepNeuralTube_{embryo}$ predicts an anterior-to-posterior shift of accessibility of those regions after the \textit{CTAAT\underline{TAG}} to \textit{CTAAT\underline{CCCT}} substitution and vice versa a posterior-to-anterior shift of accessibility after the \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} to  \textit{CTAAT\underline{TAG}} substitution (\textbf{Fig. 4f}). Next, we designed enhancers from scratch by embedding either the  \textit{CTAAT\underline{TAG}} or \textit{CTAAT\underline{CCCT}} motif\cite{taskiranCelltypedirectedDesignSynthetic2024, kempynckCREstedModelingGenomic2025} and using respectively Topic 61 and Topic 31 as target class (\textbf{Fig. 4g}). [SWITCH PANELS]. Embedding of two instances of either motif was sufficient to reach predicted chromatin accessibility levels of equivalent ranges to the predicted chromatin accessibility levels of genomic regions (\textbf{Fig. 4g}). Next, to eliminate potential model bias we repeated the same experiment but optimized for the opposite class. This is, we optimized for Topic 31 by embedding the \textit{CTAAT\underline{TAG}} motif optimized for Topic 61 by embedding the \textit{CTAAT\underline{CCCT}} motif (\textbf{Fig. 4h}). In this case, the synthetic design failed to converge for the target class but instead regions with high prediction score for the AP domain that corresponds to the motif used for embedding were generated. This is embedding of the \textit{CTAAT\underline{TAG}} motif still generated regions with high prediction score for Topic 61 and embedding of the \textit{CTAAT\underline{CCCT}} motif generated regions with high prediction score for Topic 31 (\textbf{Fig. 4h}).\par

@@

Organoid Topic 62 and embryo Topic 103 represent the pre-migratory neural crest state and instances of TEAD, GRHL, AP2 and ZEB patterns are important for this state (\textbf{Fig. 5b-c}) the importance of TEAD instances correlates with \textit{TEAD4} expression, GRHL with both \textit{GRHL1} and \textit{GRHL2} and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig. 5d}). Note that ZEB instances have a positive contribution to the pre-migratory while it has a negative contribution to all other neural crest states (\textbf{Fig. 5b-c}) and moreover all other states of both the neuronal progenitors (\textbf{Fig. 3e}) and neurons (\textbf{Fig. 6b}). The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig. 5d}). This is in line with the previously reported role of ZEB2 as a chromatin closing repressor \cite{taskiranCelltypedirectedDesignSynthetic2024}.\par

The migratory neural crest cell state is represented by two organoid topics (Topic 65 and 59) and a single embryo topic (Topic 94). Nuclear receptor, AP2, ZIC (only in organoid topics) and SOX-dimer (organoid Topic 59 but not 65) are important for this state (\textbf{Fig. 5b-c}). We could not link the nuclear receptor instances to a specific TF based on the correlation of expression and the importance score of these instances. The importance of AP2 instances correlates with the expression of \textit{TFAP2A}, \textit{TFAP2B} and \textit{TFAP2C} (\textbf{Fig. 5d}), ZIC with \textit{ZIC1} to \textit{ZIC5} and the SOX-dimer with expression of \textit{SOX10}. The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothsteinHeterodimerizationTFAP2Pioneer2020}. After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) for which instances of RFX are important (\textbf{Fig. 5b-c}) that correlate with the expression of \textit{RFX4} (\textbf{Fig. 5d}).\par

Finally, the facial mesenchyme cell state is represented by organoid Topic 58 and embryo Topic 91. Instances of a FOX
pattern and a dimer consisting of a homeobox and ebox motif are important for this state (\textbf{Fig. 5b}). The
importance of the FOX instances correlates with the expression of \textit{FOXC2} and that of the dimer motif with the
expression of both \textit{TWIST1} and \textit{ALX4}. The latter is consistent with a recent study where it has been
shown that TWIST1 and ALX4 cooperatively bind this specific dimer motif regulating the expression of genes involved in
facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024}.\par

@@

Recently, a genome wide association study (GWAS) was published identifying 203 single-nucleotide polymorphisms (SNPs)
that are associated with facial variation\cite{whiteInsightsGeneticArchitecture2021}. Given that regions surrounding these SNPs were
enriched for enhancer activity in cranial neural crest cells\cite{whiteInsightsGeneticArchitecture2021} we asked if, using
$DeepNeuralTube$, we could explain the effect of some of these variants on enhancer activity. Therefore for all
suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ and calculated the difference in prediction score across both alleles (delta prediction score). At an absolute delta prediction score threshold of 0.5 we identified X SNPs that we deem explainable by either of the two models. Note that only very few (X out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig. 5f}). Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (SUPPLEMENTAL FIG). As an example we show the explanation of rs1555067 which is located upstream of \textit{PRRX1}. The G>A substitution of this SNP causes a delta prediction score of 0.X for organoid Topic 58 (\textbf{Fig. 5g}), increasing the predicted chromatin accessibility of this genomic region in the facial mesenchyme cell state. This predicted increase in chromatin accessibility is caused by generating an EBOX site 6bp upsteam of a partial homeobox site potentially creating a new binding site for TWIST1-ALX4 dimers (\textbf{Fig. 5h}).\par

@@

\subsection*{
  Figure 5. Deciphering the enhancer code underlying pre-migratory and migratory neural crest and facial mesenchyme.
}
\textbf{a}, UMAP of organoid (top two) and embryo (bottom two) neural crest cells colored by cell-topic probabilities in RGB color scales for neural crest related topics. \textbf{b}, Code table showing average contribution score of pattern instances for each neural crest related topic. Contribution scores for $DeepNeuralTube_{organoid}$ are not outlined. Contribution scores for $DeepNeuralTube_{embryo}$ are outlined and negated. \textbf{c}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) of neural crest related topics. \textbf{d}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) for neural crest related topics. \textbf{e}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. \textbf{f}, Manhattan plot showing -log p value of the SNP being associated with variations in facial shape\cite{whiteInsightsGeneticArchitecture2021} (positive y-axis) and the maximum (across classes of the model) of the absolute value of the difference in prediction score between the reference and alternative allele (negative y-axis; circles for $DeepNeuralTube_{organoid}$ and crosses for $DeepNeuralTube_{embryo}$) across the human genome. Lead SNPs are shown with a bigger dot size. \textbf{g}, Zoom in of chromosome one with the locus of SNP rs1555067 shown as an inset, the SNP itself is shown in red. \textbf{h}, Nucleotide contribution scores for Topic 58 of $DeepNeuralTube_{organoid}$ of the references (top) and alternative (bottom) allele of SNP rs1555067. The location of the SNP is indicated with a dotted line.

@@

Summarizing our results, we identified a total of sixteen and fifteen distinct cell states in respectively the neural
tube organoids and four weeks post conception embryo head with almost one-to-one correspondence based on TFBS importances and TF expression. Additionally, we identified 4 distinct cell states corresponding to AP patterning of the neuronal progenitors in the embryo. To get insight into the minimal set of TFBS that are needed to encode these cell states we trained logistic regression models to classify genomic regions as to belonging to the topic representing each state or not, making use of the motif scores of the identified patterns from each code table (\textbf{Fig. 3e}, \textbf{Fig. 4e}, \textbf{Fig. 5b} and \textbf{Fig. 6b}). The overall performance of these models was acceptably high (\textbf{Fig. 7a}) with an average area under the ROC of X and area under the precision-recall curve of X. Notably, all models for neural crest related states performed best (\textbf{Fig. 7a}). Next, we selected patterns that had a positive coefficient for at least one of the logistic regression models resulting in a total of twenty one patterns for fifteen to sixteen states plus the additional four AP states (\textbf{Fig. 7a}). Finally, we used the identified patterns to design synthetic enhancers for each state using motif embedding\cite{taskiranCelltypedirectedDesignSynthetic2024, kempynckCREstedModelingGenomic2025}. We used at least three TFBS in the design process for each state. In case less than three distinct patterns had a positive coefficient for a state, we implanted multiple copies of each pattern to come to a total of at least four TFBS. The resulting synthetic DNA sequences had an overall high prediction score (X.X on average) (\textbf{Fig. 7a}) - and were cell type specific (\textbf{Fig. 7b}) - with the prediction score for neural crest related states being again the highest. Note that some TFBS like FOX and SOX are used by many cell states (\textbf{Fig. 7a}) while other are more specific to certain states. \par

@@

\subsection*{
  Figure 7. Neural tube cell states are encoded using four to five distinct transcription factor binding
  sites
}.
\textbf{a}, Area under the receiver-operator curve (outlined dots) and precision-recall curve of logistic regression
models classifying genomic regions to each topic using motif scores as features (top) and patterns with positive
coefficient for each logistic regression model along with the number of each patterns used for enhancer design
represented by the size of each dot (middle) and boxplots (n = 200) showing prediction score of designed enhancers for
each topic (bottom). \textbf{b}, heatmap showing prediction score for each topic of enhancers designed for all topics
for using $DeepNeuralTube_{organoid}$ (left) and $DeepNeuralTube_{embryo}$ (right). \textbf{c}, scatter plot showing
location of all pattern instances and contribution score at those locations, color is -log p value of the motif score
for each pattern calculated using FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025} for organoid (top left) and embryo
(top right) and histograms showing the number of motif instances per 25 bp bins with confidence intervals indicated in
gray scales (black is +/- 1 standard deviation, gray is +/- 2 standard deviation and white is more than 3 standard
deviations) for organoid (bottom left) and embryo (bottom right).

@@
